---
title: "Babette practical: solution"
author: "Thijs Janzen"
date: "`r Sys.Date()`"
output:
  pdf_document: default
html_document: default
editor_options:
  markdown:
  wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(babette)
library(ape)
library(beautier)
library(mauricer)
library(tracerer)
```

### Loading data

We can load data into R using the following functions, where we load an
example file from the BEAST2 examples (alternatively, download the file
from Brightspace). We then use the package ape to read the example file.

```{r loading data}
nexus_filename <- beastier::get_beast2_example_filename("Primates.nex")
fasta_filename <- "primates.fas"

nexus_data <- ape::read.nexus.data(nexus_filename)
output_data <- ape::as.DNAbin(nexus_data)

ape::write.FASTA(output_data, file = fasta_filename)
str(nexus_data)
```

The last command shows us that the dataset we have obtained containes
sequences of 12 primate species.

### Setting up babette analysis

Again, we choose a birth-death prior:

```{r tree_prior}
bd_prior <- beautier::create_tree_prior_bd()
```

#### Clock model

Again, we choose a relaxed log-normal model:

```{r clock_model}
rln_clock <- beautier::create_clock_model_rln()
```

#### Substitution model

```{r substitution model}
sub_model <- beautier::create_site_model_jc69()
```

#### Define MCMC

I choose a bit longer chain, as this tends to yield better results. Please be
aware that this may take quite some time to run!

```{r chain_setup}
mcmc_settings <- beautier::create_mcmc(chain_length = 3e6, store_every = 5000)
```

#### MRCA prior

According to Finstermeier et al 2013 (doi:10.1371/journal.pone.0069504), the 
crown age of primates is 66-69 MYA. This corresponds to a normal distribution
with mean 67.5 and sd of 0.75 (giving a 95% CI of [66, 69]). This contrasts 
later findings by Dos Reis et al. 2018 (doi:10.1093/sysbio/syy001), who compare
different clock models and obtain an estimate of 70-79 MYA (mean = 74.6, 
sd = 2.35). Which one do you think will provide better results?

```{r mrca_prior}
mrca_prior <- beautier::create_mrca_prior(
     taxa_names = beautier::get_taxa_names("primates.fas"),
     is_monophyletic = TRUE,
     mrca_distr = beautier::create_normal_distr(mean = 67.5, sigma = 0.75))
```


## Running Babette

Now we can combine our priors and mcmc settings into an inference model,
and pass this inference model to babette to start the inference

```{r inf_model}
beauti_options <- beautier::create_beauti_options_v2_6()

inf_model <- beautier::create_inference_model(tree_prior = bd_prior,
                                              clock_model = rln_clock,
                                              site_model = sub_model,
                                              mcmc = mcmc_settings,
                                              mrca_prior = mrca_prior,
                                              beauti_options = beauti_options)

beast_result <- babette::bbt_run_from_model(fasta_filename = "primates.fas",
                                            inference_model = inf_model,
                                            beast2_options = 
                                beastier::create_beast2_options(rng_seed = 421))
```

### Analysing results

Now that we have our results, we can plot them and see how we did:

```{r plot_things}
beast_result$estimates %>%
  mutate("Sample" = Sample / 1000000) %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free") +
  xlab("Sampled trees (Millions)")
```

Clearly, burnin is an issue, let's first remove 10%:

```{r remove burnin}
estimates <- tracerer::remove_burn_ins(beast_result$estimates, 0.1)
estimates %>%
  mutate("Sample" = Sample / 1000000) %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free") +
  xlab("Sampled trees (Millions)")

tracerer::calc_esses(estimates, sample_interval = 5000)
```

All ESS values are >> 200, which is very satisfying, so 10% is enough burnin.
We now also have to remove the accompanying trees of those 10% by hand:

```{r trees}
all_trees <- beast_result$primates_trees
start <- floor(0.1 * length(all_trees))
end <- length(all_trees)
all_trees <- all_trees[start:end]
```

## Resulting tree(s)

Having removed the burnin, we can now plot our results in two ways: 1)
using a densi-tree, which superimposes all trees on top of each other
and 2) using a consensus tree, which calculates the 'average' resulting
tree from our distribution of trees.

```{r plotting}
babette::plot_densitree(all_trees, alpha = 0.1)

cons_tree <- phytools::consensus.edges(trees = all_trees)
plot(cons_tree)
```

Does this tree make sense to you?

#### Age of Homo Sapiens

An interesting tidbit would be to infer the age of Homo Sapiens. Let's take
the consensus tree and find this. In here, we have to combine some information.
First, we have to know which index is used to index Homo sapiens. In the
tip.labels, we find the index of 2:
```{r print_tip}
cons_tree$tip.label
human_index <- which(cons_tree$tip.label == "Homo_sapiens")
```

Phylogenies in R are stored with two information objects: one edge table, and
one edge-length vector. The edge table indicates which nodes are connected with
each other and with tips. The edge-length vector indicates the lengths of these
connections (e.g. branch lengths). We now know to look for an edge leading 
towards tip '2':
```{r find_edge_index}
cons_tree$edge
tip_index <- which(cons_tree$edge[, 2] == human_index)
tip_index
```

This is on the 7th position. We can now use this to look up our branch length:
```{r find_age}
cons_tree$edge.length[tip_index]
```

Alternatively, we can also plot the phylogeny with edge labels!
```{r plot_edge_labels}
plot(cons_tree)
ape::edgelabels(round(cons_tree$edge.length, 2), col = "blue")
```

This informs us we share an ancestor with Pan (Chimpansee) about 12.5MYA, and an
ancestor with Gorilla ~17.3 MYA. 

If we want the uncertainty in this estimate, we have to go over all trees in
the posterior. And we can do so with lapply, which applies a function to a list.
Because our all_trees object is a list in R, we can do the following:
```{r estimate_posterior}
get_homo_age <- function(tree) {
  human_index <- which(tree$tip.label == "Homo_sapiens")
  tip_index <- which(tree$edge[, 2] == human_index)
  return(cons_tree$edge.length[tip_index])
}

all_ages <- lapply(all_trees, get_homo_age)
all_ages <- unlist(all_ages)
hist(all_ages, xlim = c(10, 16))
mean(all_ages)
median(all_ages)
quantile(all_ages, c(0.025, 0.975))
```

This shows that under our current analysis, our uncertainty of the age estimate
is very low, e.g. the 95% CI interval is very narrow around our mean estimate.
We also see that the histogram looks a bit weird. This is due to a number of 
trees (out of 9000!) where the branch lengths are very different:
```{r outliers}
length(which(all_ages < 10))
length(which(all_ages > 14))
```

## Impact of substitution model

The JC69 model is not the most realistic substitution model. Instead, let's 
see if using a GTR model changes our findings. Because we have extra parameters
to estimate, we use a longer chain.

```{r site_gtr}
sub_model <- beautier::create_site_model_gtr()

mcmc_settings <- beautier::create_mcmc(chain_length = 3e7, store_every = 5000)

inf_model <- beautier::create_inference_model(tree_prior = bd_prior,
                                              clock_model = rln_clock,
                                              site_model = sub_model,
                                              mcmc = mcmc_settings,
                                              mrca_prior = mrca_prior,
                                              beauti_options = beauti_options)

beast_result_gtr <- babette::bbt_run_from_model(fasta_filename = "primates.fas",
                                            inference_model = inf_model,
                                            beast2_options = 
                              beastier::create_beast2_options(rng_seed = 4321))
```

We find the following results:

```{r plot gtr_results}
beast_result_gtr$estimates %>%
  mutate("Sample" = Sample / 1000000) %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free") +
  xlab("Sampled trees (Millions)")
```

After removal of 10% burn-in we have:


```{r plot after burnin}
estimates_gtr <- tracerer::remove_burn_ins(beast_result_gtr$estimates, 0.1)
estimates_gtr %>%
  mutate("Sample" = Sample / 1000000) %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free") +
  xlab("Sampled trees (Millions)")

tracerer::calc_esses(estimates_gtr, sample_interval = 5000)
```

This seems sufficient to continue:

```{r trees_gtr}
all_trees_gtr <- beast_result_gtr$primates_trees
start <- floor(0.1 * length(all_trees_gtr))
end <- length(all_trees_gtr)
all_trees_gtr <- all_trees_gtr[start:end]

babette::plot_densitree(all_trees_gtr, alpha = 0.1)

cons_tree_gtr <- phytools::consensus.edges(trees = all_trees_gtr)
par(mfrow = c(1, 2))
plot(cons_tree, main = "JC69", type = "cladogram")
add.scale.bar()
plot(cons_tree_gtr, main = "GTR", type = "cladogram")
add.scale.bar()
```

Again, we see that the information contained in the sequences is very
high, overpowering choices we have made about the substitution model.

